### 2015-02-26 paper/exprs.tex
final corrections

We define three kinds of expressions: integer constants, using the value at a L-value, 
and creating a borrowed reference to the value at a L-value.

@@

\subsection*{Using a L-value}
Now that we can check that an operation is safe and we can determine from the type
which operation to perform, we can specify how to check that using the value at a L-value 
is safe and track how doing so changes the shadow heap.
\newline

@@

\begin{proof}[Justification]
  Using a path either leaves the shadow heap unchanged or deinitializes only a part of it,
  neither of which should change the types of shadows or the lifetimes of loans.
  Deinitializing part of the shadow heap could invalidate existing loans,
  but the \text{can-read} and \textsc{can-move} checks should rule out those cases.
\end{proof}

@@

Using the value at a L-value simply relies on our earlier judgment.

@@

  When creating an immutable borrow, we know from \textsc{can-read} that
  no mutable loan exists on way to the path we are borrowing or in memory reachable from it.
  The shadow of the interior of the loaned path is now controlled by the immutable loan
  of the path, but since there were no mutable loans there before and it has not changed
  it should still be well-formed. Similarly, the shadows on the way to the path have no changed
  directly and they should not be affected by the addition of an immutable loan in their
  interior. The main concern is whether the addition of the new immutable loan to the bank 
  will result in a well-formed bank. We know from the validity check that the lifetime
  of the loan is within the lifetime bound of the bank. If the bank is empty or
  the lifetime of the old loan is not shorter than the lifetime of the new loan, then
  the bank will still be well-formed. We could restrict \textsc{record} to only succeed
  in this situation.

@@

\subsection*{Using a Layout}
The runtime analogue of using a L-value is using a layout.
Affine values are replaced by void and copyable values are left untouched.
It is not defined for $\void$, which prevents expression evalutation from utilizing
uninitialized memory. A program that tries to do so will get stuck.
\newline

@@

\subsection*{Expression Evaluation}
Evaluating expressions is straightforward.
The only unusual aspect is the case of using the value at a L-value.
To prevent copies of affine values, we must overwrite the only value
with the used version. If it is a copyable value, then it will remain unchanged.
If it is an affine value, then it will be replaced with $\void$.
Note that this means the heap resulting from expression evaluation is either
unchanged or a more deinitialized version of the starting heap.
Also note that the result heap may temporarily contain a leak when considered alone,
but not when considered together with the result layout.
\newline

@@

\begin{conj}[Expression Preservation]
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$, 
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{e}{\gt;\gU'}$, and $\ev{V;H}{e}{l;H'}$
  then $\tc{\gS}{l}{\gt}$, $\tc{}{H'}{\gS}$, $\tb{\ltr}{\gU'}{\gG}{L}$, and $\tc{V}{H'}{\gU'}$.
\end{conj}

@@

\begin{conj}[Expression Progress]
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, and $\tc{\gG;\ltr;L;\gU}{e}{\gt;\gU'}$
  then there exists a layout $l$ and a heap $H'$ such that $\ev{V;H}{e}{l;H'}$.
\end{conj}

>>>>

We define three kinds of expressions: integer constants, using the value at an L-value, 
and creating a borrowed reference to the value at an L-value.

@@

\subsection*{Using an L-value}
Now that we can check that an operation is safe and we can determine from the type
which operation to perform, we can specify how to check that using the value at an L-value 
is safe and track how doing so changes the shadow heap.
\newline

@@

\begin{proof}[Justification]
  Using a path either leaves the shadow heap unchanged or deinitializes only a part of it,
  neither of which should change the types of shadows or the lifetimes of loans.
  Deinitializing part of the shadow heap could invalidate existing loans,
  but the \textsc{can-read} and \textsc{can-move} checks should rule out those cases.
\end{proof}

@@

Using the value at an L-value simply relies on our earlier judgment.

@@

  When creating an immutable borrow, we know from \textsc{can-read} that
  no mutable loan exists on way to the path we are borrowing or in memory reachable from it.
  The shadow of the interior of the loaned path is now controlled by the immutable loan
  of the path, but since there were no mutable loans there before and it has not changed
  it should still be well-formed. Similarly, the shadows on the way to the path have not changed
  directly and they should not be affected by the addition of an immutable loan in their
  interior. The main concern is whether the addition of the new immutable loan to the bank 
  will result in a well-formed bank. We know from the validity check that the lifetime
  of the loan is within the lifetime bound of the bank. If the bank is empty or
  the lifetime of the old loan is not shorter than the lifetime of the new loan, then
  the bank will still be well-formed. We could restrict \textsc{record} to only succeed
  in this situation.

@@

\subsection*{Using a Layout}
The runtime analogue of using an L-value is using a layout.
Affine values are replaced by void and copyable values are left untouched.
It is not defined for $\void$, which prevents expression evalutation from utilizing
uninitialized memory. A program that tries to do so will get stuck.
\newline

@@

\subsection*{Expression Evaluation}
Evaluating expressions is straightforward.
The only unusual aspect is the case of using the value at an L-value.
To prevent copies of affine values, we must overwrite the original value
with the used version. If it is a copyable value, then it will remain unchanged.
If it is an affine value, then it will be replaced with $\void$.
Note that this means the heap resulting from expression evaluation is either
unchanged or a more deinitialized version of the starting heap.
Also note that the result heap may temporarily contain a leak when considered alone,
but not when considered together with the result layout.
\newline

@@

\begin{conj}[Expression Preservation]
  \raggedright
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$, 
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{e}{\gt;\gU'}$, and $\ev{V;H}{e}{l;H'}$
  then $\tc{\gS}{l}{\gt}$, $\tc{}{H'}{\gS}$, $\tb{\ltr}{\gU'}{\gG}{L}$, and $\tc{V}{H'}{\gU'}$.
\end{conj}

@@

\begin{conj}[Expression Progress]
  \raggedright
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, and $\tc{\gG;\ltr;L;\gU}{e}{\gt;\gU'}$
  then there exists a layout $l$ and a heap $H'$ such that $\ev{V;H}{e}{l;H'}$.
\end{conj}

### 2015-02-26 paper/futurework.tex
final corrections

Until recently, Rust's closures have been in flux. As of this writing, Rust supports
both stack and heap closures and also supports different manners of access to the environment
(owning, immutable refernce, or mutable reference). In the past, closures have been a
common source of borrow checker bugs, which likely means that proving soundness will be
difficult. However, it also means that it is one of the most essential features to add to Patina.
It would likely be profitable to look to Cyclone \cite{cycregions}, which took great
care to ensure references could not escape via closures or existential types.

>>>>

Until recently, Rust's closures have been in flux. As of this writing, Rust supports
both stack and heap closures and also supports different manners of access to the environment
(owning, immutable reference, or mutable reference). In the past, closures have been a
common source of borrow checker bugs, which likely means that proving soundness will be
difficult. However, it also means that it is one of the most essential features to add to Patina.
It would likely be profitable to look to Cyclone \cite{cycregions}, which took great
care to ensure references could not escape via closures or existential types.

### 2015-02-26 paper/intro.tex
final corrections

\section*{Abstract}
Rust is a new systems language that uses some advanced type system features,
specifically affine types and regions, to statically guarantee memory
safety and eliminate the need for a garbage collector.
While each individual addition to the type system is well understood
in isolation and are known to be sound, the combined system is not.
Furthermore, Rust uses a novel checking scheme for its regions, known
as the Borrow Checker, that is not known to be correct.
Since Rust's goal is to be a safer alternative to C/C++, we should
ensure that this safety scheme actually works. We present a formal
semantics that camptures the key features relevant to memory safety,
unique pointers and borrowed references, specifies how they guarantee
memory safety, and describes the operation of the Borrow Checker.
We used this model to prove the soudness of some core operations
and justify the conjecture that the model, as a whole, is sound.
Additionally, our model provides a syntactic version of the Borrow
Checker, which may be more understandable than the non-syntactic version
in Rust.

>>>>

\section*{Abstract}
Rust is a new systems language that uses some advanced type system features,
specifically affine types and regions, to statically guarantee memory
safety and eliminate the need for a garbage collector.
While each individual addition to the type system is well understood
in isolation and are known to be sound, the combined system is not.
Furthermore, Rust uses a novel checking scheme for its regions, known
as the Borrow Checker, that is not known to be correct.
Since Rust's goal is to be a safer alternative to C/C++, we should
ensure that this safety scheme actually works. We present a formal
semantics that captures the key features relevant to memory safety,
unique pointers and borrowed references, specifies how they guarantee
memory safety, and describes the operation of the Borrow Checker.
We use this model to prove the soudness of some core operations
and justify the conjecture that the model, as a whole, is sound.
Additionally, our model provides a syntactic version of the Borrow
Checker, which may be more understandable than the non-syntactic version
in Rust.

### 2015-02-26 paper/paths.tex
final corrections

\section*{Read Safety}
We are now in position to prove that our read operation is safe.
That is, if the heap is well formed and a route is has some type,
then we can read some layout at that route and that layout has that same type.
This is effectively a merger of the type preservation and progress lemmas for \textsc{read}.

@@

Finally, we can state and prove path type preservation.
Under well-formed heap and contexts, typed paths evaluate to routes of the same type.

@@

Since shadows and holes are supposed to be a shadow of the heap,
it is unsurprising that they are very similar to layouts and cells.
However, there are a few key differences.
First, variants are considered an atomic value (and thus a hole) rather than
a structure. This is because the discrimiant and payload of a variant always
share the same initialization state.
Second, while the pointer cells contain routes, the pointer holes do not
contain paths (the analogue of routes).
For references, there is no need because references always point to fully initialized things.
Therefore, the reference hole simply records the type of the reference.
For unique pointers, there is no path that could possibly be used
(otherwise the pointer would not be unique).
Therefore, the unique pointer hole records the shadow of the pointed-to heap memory.

@@

\section*{Shadow Typing}
We need to ensure that the shadows assigned to variables are consistent with the types
assigned to variables, i.e. we want to avoid situations like
$\gG(x)=\tyint$ and $\gU(x)=\refval{\qimm}{\tyint}$.
We can easily do this by defining a typing judgement for shadows and
defining a well-formedness condition for $\gU$ that ensures it is consistent with $\gG$.
Similar to typing for layouts, $\uninit$ can only have types that other holes can have.

@@

\section*{Shadow Checking the Heap}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot leak heap memory by doing so.
The $\tc{V}{H}{\gU}$ judgment ensures that the shadow heap corresponds to the real heap.
\newline

@@

\begin{mathpar}
\mprset{flushleft}
\infer
{ 
\dom{V}=\dom{\gU} \\\\
\dom H = \textsc{reachable}(V,H) \\\\
\forall x \in \dom{V}.~\tc{H}{H(V(x))}{\gU(x)}
}
{\tc{V}{H}{\gU}}
\end{mathpar}

>>>>

\section*{Read Safety}
We are now in position to prove that our read operation is safe.
That is, if the heap is well formed and a route has some type,
then we can read some layout at that route and that layout has that same type.
This is effectively a merger of the type preservation and progress lemmas for \textsc{read}.

@@

Finally, we can state and prove path type preservation.
Under a well-formed heap and contexts, typed paths evaluate to routes of the same type.

@@

Since shadows and holes are supposed to be a shadow of the heap,
it is unsurprising that they are very similar to layouts and cells.
However, there are a few key differences.
First, variants are considered an atomic value (and thus a hole) rather than
a structure. This is because the discrimiant and payload of a variant always
share the same initialization state.
Second, while the pointer cells contain routes, the pointer holes do not
contain paths (the analogue of routes).
For references, there is no need because references always point to fully initialized things.
Therefore, the reference hole simply records the type of the referent.
For unique pointers, there is no path that could possibly be used
(otherwise the pointer would not be unique).
Therefore, the unique pointer hole records the shadow of the pointed-to heap memory.

@@

\section*{Shadow Typing}
We need to ensure that the shadows assigned to variables are consistent with the types
assigned to variables, i.e. we want to avoid situations like
$\gG(x)=\tyint$ and $\gU(x)=\refval{\qimm}{\tyint}$.
We can easily do this by defining a typing judgement for shadows and
defining a well-formedness condition for $\gU$ that ensures it is consistent with $\gG$.
Similar to typing for layouts, $\uninit$ can only have types that other holes can have.
\newline

@@

\section*{Shadow Checking the Heap}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\ga}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot leak heap memory by doing so.
The $\tc{V}{H}{\gU}$ judgment ensures that the shadow heap corresponds to the real heap.
It also ensures that all heap memory is reachable from the stack and that
every heap allocation has at least one owner: either a stack variable or a unique pointer.
\newline

@@

\begin{mathpar}
\mprset{flushleft}
\infer
{ 
\dom{V}=\dom{\gU} \\\\
\dom H = \textsc{reachable}(V,H) \\\\
\forall \ga \in \dom H. \exists x. V(x) = \ga \lor \exists \addr{\ga'}{r}. \Read{H}{\addr{\ga'}{r}}{\own{\ga}} \\\\
\forall x \in \dom{V}.~\tc{H}{H(V(x))}{\gU(x)}
}
{\tc{V}{H}{\gU}}
\end{mathpar}

### 2015-02-26 paper/relatedwork.tex
final corrections

Gordon et al. \cite{uniqueandrefimm} introduce a static method for ensuring data-race freedom
by enforcing that at most one mutable reference to an object exists, which is a similar task
to Rust's effort to avoid aliased, mutable memory.
Their work supports read-only references, which cannot be used to mutate, but to not forbid
mutation. Since these are by definition aliased, mutable memory, Rust has no equivalent.
However, their immutable reference is similar to Rust's immutable borrowed references,
and their isolated referece is somewhat similar to Rust's unique pointers and mutable
borrowed references. Isolated references can be coverted to writeable references and back,
which allows for behavior similar to a mutable borrowed reference. Isolated references
can also be converted into immutable references, which is analogous to immutably reborrowing
a mutably borrowed reference. Their correspondence with unique pointers is similar.
Gordon, Ernst, and Grossman \cite{relyguarantee} later propose a general notion of
rely-guarantee references, which treat aliases similar to different threads of control
and use methods from rely-guarantee program logics to ensure safety. Rust's references
can likely be encoded in this general framework.

@@

Clarke and Wrigstad's external uniqueness \cite{externaluniqueness} has a notion
of ownership similar to Rust's. Both have recursive ownership; however, Rust uses
it for memory management rather than for access control.
Unlike Rust, unique pointers are nullified on use. The Patina model nullifies on use,
but soundness ensures it does not matter in practice.
There is also fresh owner notion for borrowed references that accomplishes the same
thing as Rust's lifetimes, i.e. preventing escape.
Unlike Rust, Clarke and Wrigstad's borrowed references are both copyable and mutable,
which means they cannot make the same safety guarantees as Rust, i.e. no access to
uninitialized memory.
They list several options about how borrowed unique pointers can be handled, but
Rust does not cleanly fit into these categories. Rust discriminates on the manner of borrowing
and the operation to be performed to determine whether it is safe to use the borrowed
unique pointer, but Clarke and Wrigstad suggest that a system must either allow all uses
or forbid all uses.
Their model also has problems with immutable (final) variables and unique versions
of non-nullable pointers because they denote consumption via nullification and the finality
prevents that. Rust, however, has both immutable variables and all of its pointers are non-null,
but it has no issues mixing these with uniqueness. This is because Rust's ban on null pointers
is not a ban on pointers being null, but rather on pointers being used while null.
Additionally, Rust's immutable variables' usability can change if a unique value is consumed,
which allows Rust to render an immutable variable unusable thereafter.
Clarke and Wrigstad speculate that alias burying would enable them to support this combination,
which is somewhat true. Rust does manage this via static analysis, but not via alias burying.

@@

>>>>

Gordon et al. \cite{uniqueandrefimm} introduce a static method for ensuring data-race freedom
by enforcing that at most one mutable reference to an object exists, which is a similar task
to Rust's effort to avoid aliased, mutable memory.
Their work supports read-only references, which cannot be used to mutate, but do not forbid
mutation. Since these are by definition aliased, mutable memory, Rust has no equivalent.
However, their immutable reference is similar to Rust's immutable borrowed references,
and their isolated referece is somewhat similar to Rust's unique pointers and mutable
borrowed references. Isolated references can be coverted to writeable references and back,
which allows for behavior similar to a mutable borrowed reference. Isolated references
can also be converted into immutable references, which is analogous to immutably reborrowing
a mutably borrowed reference. Their correspondence with unique pointers is similar.
Gordon, Ernst, and Grossman \cite{relyguarantee} later propose a general notion of
rely-guarantee references, which treat aliases similar to different threads of control
and use methods from rely-guarantee program logics to ensure safety. Rust's references
can likely be encoded in this general framework.

@@

Clarke and Wrigstad's external uniqueness \cite{externaluniqueness} has a notion
of ownership similar to Rust's. Both have recursive ownership; however, Rust uses
it for memory management rather than for access control.
Unlike Rust, unique pointers are nullified on use. The Patina model nullifies on use,
but soundness ensures it does not matter in practice.
There is also a fresh owner notion for borrowed references that accomplishes the same
thing as Rust's lifetimes, i.e. preventing escape.
Unlike Rust, Clarke and Wrigstad's borrowed references are both copyable and mutable,
which means they cannot make the same safety guarantees as Rust, i.e. no access to
uninitialized memory.
They list several options about how borrowed unique pointers can be handled, but
Rust does not cleanly fit into these categories. Rust discriminates on the manner of borrowing
and the operation to be performed to determine whether it is safe to use the borrowed
unique pointer, but Clarke and Wrigstad suggest that a system must either allow all uses
or forbid all uses.
Their model also has problems with immutable (final) variables and unique versions
of non-nullable pointers because they denote consumption via nullification and the finality
prevents that. Rust, however, has both immutable variables and all of its pointers are non-null,
but it has no issues mixing these with uniqueness. This is because Rust's ban on null pointers
is not a ban on pointers being null, but rather on pointers being used while null.
Additionally, Rust's immutable variables' usability can change if a unique value is consumed,
which allows Rust to render an immutable variable unusable thereafter.
Clarke and Wrigstad speculate that alias burying would enable them to support this combination,
which is somewhat true. Rust does manage this via static analysis, but not via alias burying.

@@

Potanin et al. \cite{ooimmutability} survey the various ways of adding immutability
to object-oriented languages and how it impacts aliasing, including Javari, OIGJ, and Joe$_3$.
These systems usually add immutable objects and/or read-only references, but they do not
provide the no aliased, mutable memory guarantee of Rust. As discussed earlier, read-only
references are antithetical to Rust's memory safety invariant. Immutable objects are similar
to Rust's immutably borrowed references, but the OO systems often go further by allowing
immutability at the field level. Rust does not allow programmers to specify the
mutability of particular fields, but immutable borrows can provide a temporary guarantee
that is similar. Several of the OO systems have a concept of ownership, but it is primarily
about encapsulation and access control. 
Rust handles encapsulation via its module system rather than
via the type system or static analysis. As usual with OO languages, they rely on garbage
collection for memory management, which is very unlike Rust.

### 2015-02-26 paper/stmts.tex
final corrections

\section*{Statements}
We have finally reached the outermost shell of Patina!
The statements layer is primarily about heap manipulation and chaining expressions.
It mostly just relies on judgments we have already defined.
$\nop$ and $\seq{s}{s}$ are simply for chaining together statements.
$\set{\lv{x}{p}}{e}$ is for initializing a path with the result of an expression.
$\new{\lv{x}{p}}{\lv{x}{p}}$ is similar, but allocates the result on the heap 
and initializes the path with a new unique pointer. 
Treating this as a statement rather than an expression
simplifies the expression preservation and progress lemmas.
Note that both versions are for \emph{initialization} and not \emph{assignment}.
We only allow initialization of droppable paths, which we guarantee will not leak memory.
$\swap{\lv{x}{p}}{\lv{x}{p}}$ will perform a shallow swap of two writable paths.
This is the only way to actually mutate through a mutable borrow since we do not allow
interior of a borrowed reference to be uninitialized even temporarily.
$\free{\lv{x}{p}}$ performs a shallow free and is the inverse of \textsf{new}.
$\push{\lt}{x}{\gt}{s}$ will create a new stack variable.
$\pop{x}{s}$ will pop that stack variable once the statement that requires it finishes.
It is for evaluation purposes and should not appear in user code.

@@

\subsection*{Statement Evaluation}
Statement evaluation goes as you would expect.
We use a new helper judgment $\textsc{fresh}(\ga)$ that generates a unused allocation.
As before, we utilize domain restriction to "deallocate".
Other than that, we simply use path and expression evaluation combined
with \textsc{read} and \textsc{write}.
\newline

@@

It is difficult to specify how the new versions of contexts relate to the old versions.
They may be larger or smaller depending upon the statement in question.
We do not specify this relationship currently.
Instead we just rely on the new variable map, heap and statement to enforce
the relationship.

\begin{conj}[Statement Preservation]
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{s}{\gU_0}$, and $\sev{V}{H}{s}{V'}{H'}{s'}$
  then there exist $\gS'$, $\gG'$, $\ltr'$, $L'$, and $\gU'$ such that
  $\tc{}{H'}{\gS'}$, $\gG';V'\vdash\gS'$, $\vdash\ltr'$, $\tb{\ltr'}{\gU'}{\gG'}{L'}$,
  $\tc{V'}{H'}{\gU'}$, and $\tc{\gG';\ltr';L';\gU'}{s'}{\gU_0}$.

@@

\section*{Patina Soundness}
If our conjectures hold, then we have established soundness for Patina.
This means that well-typed statement that ends with an empty shadow heap
will not leak memory, not access uninitialized memory, not create aliased, mutable memory,
and so on. In effect, well-typed programs are statically guaranteed to be memory safe.

>>>>

\section*{Statements}
We have finally reached the outermost shell of Patina!
The statements layer is primarily about heap manipulation and chaining expressions.
It mostly just relies on judgments we have already defined.
$\nop$ and $\seq{s}{s}$ are simply for chaining together statements.
$\set{\lv{x}{p}}{e}$ is for initializing a path with the result of an expression.
$\new{\lv{x}{p}}{\lv{x}{p}}$ is similar, but allocates the result on the heap 
and initializes the path with a new unique pointer. 
Treating this as a statement rather than an expression
simplifies the expression preservation and progress lemmas.
Note that both versions are for \emph{initialization} and not \emph{assignment}.
We only allow initialization of droppable paths, which we guarantee will not leak memory.
$\swap{\lv{x}{p}}{\lv{x}{p}}$ will perform a shallow swap of two writable paths.
This is the only way to actually mutate through a mutable borrow since we do not allow
the interior of a borrowed reference to be uninitialized even temporarily.
$\free{\lv{x}{p}}$ performs a shallow free and is the inverse of \textsf{new}.
$\push{\lt}{x}{\gt}{s}$ will create a new stack variable.
$\pop{x}{s}$ will pop that stack variable once the statement that requires it finishes.
It is for evaluation purposes and should not appear in user code.

@@

\subsection*{Statement Evaluation}
Statement evaluation goes as you would expect.
We use a new helper judgment $\textsc{fresh}(\ga)$ that generates a unused allocation.
As before, we utilize domain restriction to ``deallocate''.
Other than that, we simply use path and expression evaluation combined
with \textsc{read} and \textsc{write}.
\newline

@@

It is difficult to specify how the new versions of contexts relate to the old versions.
They may be larger or smaller depending upon the statement in question.
We do not specify this relationship currently.
Instead we just rely on the new variable map, heap, and statement to enforce
the relationship.

\begin{conj}[Statement Preservation]
  \raggedright
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{s}{\gU_0}$, and $\sev{V}{H}{s}{V'}{H'}{s'}$
  then there exist $\gS'$, $\gG'$, $\ltr'$, $L'$, and $\gU'$ such that
  $\tc{}{H'}{\gS'}$, $\gG';V'\vdash\gS'$, $\vdash\ltr'$, $\tb{\ltr'}{\gU'}{\gG'}{L'}$,
  $\tc{V'}{H'}{\gU'}$, and $\tc{\gG';\ltr';L';\gU'}{s'}{\gU_0}$.

@@

\section*{Patina Soundness}
If our conjectures hold, then we have established soundness for Patina.
This means that a well-typed statement that ends with an empty shadow heap
will not leak memory, not access uninitialized memory, not create aliased, mutable memory,
and so on. In effect, well-typed programs are statically guaranteed to be memory safe.
